Type Evolution in Programming Systems

### 2024-05-29 paper.tex
\openframeworkbox \closeframeworkbox to handle math

\DeclareRobustCommand{\frameworkbox}[2][gray!15]{
\begin{tcolorbox}[breakable,left=3pt,right=3pt,top=3pt,bottom=3pt,colback=#1,colframe=#1,parbox=false,
  width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
  outer arc=0pt]\setlength{\parskip}{0.5em}#2
\end{tcolorbox}}
\newcommand{\frameworkboxtitle}[1]{\setlength{\parindent}{0em}{\firamedium #1.}\quad}

@@

\begin{figure}[H]
\begin{tikzcd}[sep = 6em]
  \textsf{type}_1
  \arrow[r, "\textsf{constrains}"{above}]
  \arrow[d, red, "\textsf{change}_1"{name=C, description}]
  &
  \textsf{data}_1
  \arrow[d, "\textsf{migrate}_1"{name=R, description}]
  \\
  \textsf{type}_2
  \arrow[r, "\textsf{constrains}"{above}]
  \arrow[d, red, "\textsf{change}_2"{name=C2, description}]
  &
  \textsf{data}_2
  \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
  \arrow[d, "\textsf{migrate}_2"{name=R2, description}]
  \\
  \textsf{type}_3
  \arrow[r, "\textsf{constrains}"{above}]
  &
  \textsf{data}_3
  \arrow[Rightarrow, from=C2, to=R2, "\textsf{propagate}"{above}]
\end{tikzcd}
\caption{Multiversion type evolution.}
\label{fig:multiversion}
\end{figure}

\frameworkbox{
% wrapfigure doesn't seem to work with tikz
% \begin{wrapfigure}{r}{9em}
% \includegraphics[scale=0.2]{figures/diag-variants.png}
% \end{wrapfigure}

\frameworkboxtitle{Primitive Type Transformations}
In this example, \textsf{type} in Figure~\ref{fig:multiversion} refers to data tables, but the same scenario applies to
objects or JSON values in other programming systems. Our example focuses on permanent \emph{data}
that need to evolve alongside with the type and keep synchronized even in the non-local scenario.
The evolution can be done through a small set of primitive type transformations (illustrated on the
right) with corresponding data transformation for each. The system may either expose the
primitive transformations through a user interface, or infer them from textual edits made by the
programmer, e.g. to a database table schema definition.
}

>>>>

\DeclareRobustCommand{\frameworkbox}[2][gray!15]{
\begin{tcolorbox}[breakable,left=3pt,right=3pt,top=3pt,bottom=3pt,colback=#1,colframe=#1,parbox=false,
  width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
  outer arc=0pt]\setlength{\parskip}{0.5em}#2
\end{tcolorbox}}
\newcommand{\frameworkboxtitle}[1]{\setlength{\parindent}{0em}{\firamedium #1.}\quad}
\newcommand{\openframeworkbox}{
\begin{tcolorbox}[breakable,left=3pt,right=3pt,top=3pt,bottom=3pt,colback=gray!15,colframe=gray!15,parbox=false,
  width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
  outer arc=0pt]\setlength{\parskip}{0.5em}
}
\newcommand{\closeframeworkbox}{
\end{tcolorbox}
}

@@

\openframeworkbox
  \begin{tikzcd}[sep = 6em]
    \textsf{type}_1
    \arrow[r, "\textsf{constrains}"{above}]
    \arrow[d, red, "\textsf{change}_1"{name=C, description}]
    &
    \textsf{data}_1
    \arrow[d, "\textsf{migrate}_1"{name=R, description}]
    \\
    \textsf{type}_2
    \arrow[r, "\textsf{constrains}"{above}]
    \arrow[d, red, "\textsf{change}_2"{name=C2, description}]
    &
    \textsf{data}_2
    \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
    \arrow[d, "\textsf{migrate}_2"{name=R2, description}]
    \\
    \textsf{type}_3
    \arrow[r, "\textsf{constrains}"{above}]
    &
    \textsf{data}_3
    \arrow[Rightarrow, from=C2, to=R2, "\textsf{propagate}"{above}]
  \end{tikzcd}

  \frameworkboxtitle{Primitive Type Transformations}
In this example, \textsf{type} refers to data tables, but the same scenario applies to
objects or JSON values in other programming systems. Our example focuses on permanent \emph{data}
that need to evolve alongside with the type and keep synchronized even in the non-local scenario.
The evolution can be done through a small set of primitive type transformations (illustrated on the
right) with corresponding data transformation for each. The system may either expose the
primitive transformations through a user interface, or infer them from textual edits made by the
programmer, e.g. to a database table schema definition.
\closeframeworkbox
